---
title: "run_biomee_with_LUH2"
output: rmarkdown::html_vignette
---

## Overview

BiomeE is a dynamic vegetation demography simulation engine capable of predicting the evolution of a *vegetation tile* characterized by static parameters (species, soil parameters, initial vegetation, etc)
when subject to environmental forcing and land use change forcing. To learn more, please refer to `vignette("BiomeE usage", package = "rsofun")` and `vignette("BiomeE LULUC", package = "rsofun")`. 

## LUH2 integration

BiomeE integrates seamlessly with LUH2 v2 data (https://luh.umd.edu/data.shtml).

The rsofun package contains an example of driver from the Swiss CH-LAE fluxnet site, using environmental data for the year 2009,
where 40% of an initial primary forest is clear-cut and convert to a secondary (yet unmanaged) forest.

```{r eval = TRUE}
library(rsofun)
library(tidyr)

rsofun::biomee_p_model_luluc_drivers
```

### Automatic configuration: TODO: remove this functionality, always do manual configuration

Below is an example of minimal configuration:

```{r eval = FALSE}
df_drivers <- biomee_p_model_drivers
# Disable daily diagnostics (optional)
df_drivers$params_siml[[1]]$daily_diagnostics <- FALSE

# Simulate for 10 years after the spinup period. By default the same number of years are extracted from LUH2 dtaa.
df_drivers$params_siml[[1]]$nyeartrend <- 10
# In order to extract a custom number of years, set the following:
# df_drivers$luh2[[1]]$n <- ...
# By default the extraction start at the first year in the ncdf files. To extract from a custom index:
# df_drivers$luh2[[1]]$start <- ...

# Setting the paths to LUH2 files. These may be different on your computer.
df_drivers$luh2[[1]]$state_file <- 'path/to/states.nc'
df_drivers$luh2[[1]]$trans_file <- 'path/to/transitions.nc'

# Indicate whether to use simplified states.
df_drivers$luh2[[1]]$simplified <- TRUE
# By default all the states defined in LUH2 v2 are used (12 states)
# In the simplified mode, only 5 states are used:
# prim     <- primf + primn
# secd     <- secdf + secdn
# urban    <- urban
# cropland <- cxxxx
# pasture  <- pastr + range
```

Then, run the simulation as before.

Notes:

- Extracting data from LUH2 ncdf files may take some time.
- The cell that is extracted from LUH2 data is the closest cell from the coordinates provided in `site_info`.
- If no `init_lu` is present in the configuration, a default configuration is automatically generated. If you wish to provide your own, make sure to use the right state names.
- When using LUH2 data, `luc_forcing` must not be provided in the driver.
- The fraction of the aggregated tile should always remain constant. In most cases, it would be close to 1. But it may be less in cases where the cell contains a non-null water/ice fraction. This fraction is defined in `staticData_quarterdeg.nc/icwtr`.
- BiomeE assumes a cell of arbitrary area (i.e. the cell has an area of 1). To convert densities into real units, all densities should be scaled by the area of the cell, which is available in `staticData_quarterdeg.nc/carea`.

### Manual configuration

The automatic configuration presented above is convenient but has some drawbacks, such as the need to parse LUH2 data each time a simulation is run.
An alternative offering more flexibility is to parse LUH2 manually and set the necessary data in the driver before running BiomeE.
This can easily be done using `parse_luh2()` function as illustrated below:

```{r, eval = FALSE}

# We start from the example driver
df_drivers_luh2 <- biomee_p_model_luluc_drivers

# LUH2 v2f dataset contains 85 years (2015-2100)
df_drivers_luh2$params_siml[[1]]$nyeartrend <- 85
df_drivers_luh2$params_siml[[1]]$daily_diagnostics <- FALSE

# Data v2f SSP1 RPC2.6 downloaded from https://luh.umd.edu/data.shtml)
source("../data-raw/generate_LUH2_data.R") # downloads the LUH2 data and returns paths `state_file` and `trans_file`
# state_file <- "/tmp/Rtmpn6fjSh/multiple-states_input4MIPs_landState_ScenarioMIP_UofMD-IMAGE-ssp126-2-1-e_gn_2100-2300.nc"
# trans_file <- "/tmp/Rtmpn6fjSh/multiple-transitions_input4MIPs_landState_ScenarioMIP_UofMD-IMAGE-ssp126-2-1-e_gn_2100-2300.nc"

# Parse LUH2 data
parsed_luh2 <- parse_luh2(
  state_file = state_file, 
  trans_file = trans_file,
  lon        = df_drivers_luh2$site_info[[1]]$lon, 
  lat        = df_drivers_luh2$site_info[[1]]$lat,
  start      = 1, 
  n          = df_drivers_luh2$params_siml[[1]]$nyeartrend, # We parse the whole dataset
  simplified = TRUE # simplified mode: 5 states only
)

# Create the init_lu config
init_lu <- tibble(
  # Keep these two lines
  name      = names(parsed_luh2$states_init),
  fraction  = parsed_luh2$states_init,
  # The following line should be tailored to your application:
  preset    = c(rep('unmanaged', 2), 'urban', 'cropland', 'pasture')
  # In addition more fine-grained parameters could be set ('extra_N_input', 'extra_turnover_rate', 'oxidized_litter_fraction')
)

# TODO: check_luc_matrix(parsed_luh2$luc_matrix, init_lu)
# TODO: luc_matrix_to_luc_df(parsed_luh2$luc_matrix, init_lu$name)
# TODO: plot_luc_matrix(parsed_luh2$luc_matrix, init_lu)

# Set init_lu and luc_forcing in driver
df_drivers_luh2$init_lu[[1]]     <- init_lu
df_drivers_luh2$luc_forcing[[1]] <- parsed_luh2$luc_matrix

# Note: in this example, we do not set df_drivers$init_cohort for brevity, but it should be done as well as demonstrated in the previous sections.

```

```{r load luluc driver, eval = TRUE, include = FALSE}
# Since the LUH2 data is not stored in this repo, we load the driver from disk
#saveRDS(df_drivers_luh2, "files/biomee_luluc.Rmd__biomee_p_model_luluc_driver.RDS")
df_drivers_luh2 <- readRDS("files/biomee_luluc.Rmd__biomee_p_model_luluc_driver.RDS")
```

We can now simulate and display the plots as before:

```{r eval = TRUE, results = 'hide', warning = FALSE, fig.width=7, fig.height=7}
out_sim <- runread_biomee_f(df_drivers_luh2)
```

Next, we define some convenience functions for displaying plots:

```{r eval = TRUE, results = 'hide'}
library(ggplot2)
library(gridExtra)
library(purrr)

plot1 <- function(lu_name, variable, out, y_limit=NA, yr_start=1, yr_label_offset=0) {
  tile <- out[[lu_name]][[1]]
  if(lu_name != 'aggregated'){
    tile <- tile$output_annual_tile
  } else {
    tile <- tile$output_annual_cell
  }
          
  if (variable != 'lu_fraction') {
    res <- tile %>%
          ggplot() +
          geom_line(aes(x = year + yr_label_offset,
                        y = get(variable) * lu_fraction)) +
          coord_cartesian(xlim = c(yr_start, NA),
                          ylim = c(0, y_limit)) +
          theme_classic() +
          labs(x = "Year", y = paste(variable, "(", lu_name, ")"))
  } else {
    stopifnot(variable == 'lu_fraction')
    res <- tile %>%
          ggplot() +
          geom_line(aes(x = year + yr_label_offset,
                        y = get(variable))) +
          coord_cartesian(xlim = c(yr_start, NA),
                          ylim = c(0, y_limit)) +
          theme_classic() +
          labs(x = "Year", y = paste(variable, "(", lu_name, ")"))
  }

  return(res)
}

plot_variable <- function(variable, out, yr_start=1, yr_label_offset=0) {
  agg <- out[['aggregated']][[1]]$output_annual_cell
  y_limit <- max(agg[variable] * agg$lu_fraction) * 1.01

  # We remove sitename and aggregated
  tile_names <- names(out)[3:length(names(out))]

  names <- c(tile_names, 'aggregated')

  names |> lmap(\(x) plot1(x, variable, out, y_limit, yr_start, yr_label_offset))
}

plot_variables <- function(variables, out, yr_start=1, yr_label_offset=0) {

  plots <- variables |> lmap(\(x) plot_variable(x, out, yr_start, yr_label_offset))
  grid.arrange(grobs=plots, nrow = length(variables))
}
```

``` {r}
plot_variables(c('GPP', 'fastSOM', 'lu_fraction'), 
               out_sim |> dplyr::select(-prim), 
               1900, 
               2015 - df_drivers_luh2$params_siml[[1]]$spinupyears - 1)
```
